---
title: "Browsing the spatial data catalogue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Browsing the spatial data catalogue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Chunks are not evaluated: the catalogue reads the ABMI science
# share, which is not reachable when the package is built or
# checked. Output blocks below are transcripts from a real scan.
# The one exception is the catalogue table, which is rendered from
# a frozen scan committed alongside this file
# (data-raw/make_catalogue_snapshot.R writes it).
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

```{r}
library(sciSpatialR)
```

## The share is the catalogue

There is no manifest file to maintain. `build_catalogue()` walks
`\\ABMI-DATA2\science\spatial_data`, parses the readme stored beside
each dataset, and assembles a table from what it finds. Documenting a
dataset means editing its readme; nothing has to be registered
anywhere, and the catalogue cannot drift from the data.

Folders follow the ISO 19115 topic categories and readmes follow the
ABMI spatial metadata template, both documented in the
[geospatial catalog and management guide](https://github.com/bgcasey/geospatial_catalog_and_management_guide).

A folder is catalogued as a **layer** when it holds a readme written
to the product or variant template. Theme folders carry a shorter
`Category`/`Description`/`Examples` readme and are reported by
`list_themes()` instead. Products sit directly under their topic
category, so each layer has both a short `name` (`fab_dem`) and a full
`id` (`elevation/fab_dem`). A product may split its record across a
product readme and one readme per variant subfolder, in which case the
variant is the layer and its `id` carries the variant folder too
(`geoscientificInformation/soilgrids_250_v2_ab/abmi1km`).

The one folder no scan looks at is top-level `_temp`, the share's
scratch area. Work in progress, staging copies, and exports live
there, so its contents are neither catalogued as layers nor counted
against the share by `check_metadata()` — a layer only becomes
discoverable once it is filed under a theme.

## Pointing at the share

`spatial_root()` reports where the catalogue reads from and errors if
the share is unreachable.

```{r}
spatial_root()
#> [1] "//ABMI-DATA2/science/spatial_data"
```

To work from a mirror, a mapped drive, or a local copy, set an option
or an environment variable:

```{r}
options(sciSpatialR.spatial_root = "D:/spatial_data")
Sys.setenv(SCISPATIALR_SPATIAL_ROOT = "D:/spatial_data")
```

## Browsing

Start with the themes, which show the topic categories and how much
sits under each:

```{r}
list_themes()
```

```
14 themes, 20 layers catalogued

 theme                            description                                        n_layers
 elevation                        Height above or below sea level.                   5       
 inlandWaters                     Inland water features, drainage systems, and thei… 4       
 imageryBaseMapsEarthCover        Base maps.                                         3       
 biota                            Flora and/or fauna in natural environments.        2       
 geoscientificInformation         Information pertaining to earth sciences.          2       
 location                         Positional information and services.               2       
 boundaries                       Legal land descriptions.                           1       
 transportation                   Means and aids for conveying persons and/or goods. 1       
 .git                             <NA>                                               0       
 climatologyMeteorologyAtmosphere Processes and phenomena of the atmosphere.         0       
 economy                          Economic activities, conditions, and employment.   0       
 environment                      Environmental resources, protection, and conserva… 0       
 farming                          Rearing of animals and/or cultivation of plants.   0       
 figures                          <NA>                                               0       
```

Every ISO 19115 topic category appears, including the ones nothing is
filed under yet, so the empty folders read as available rather than
missing. A theme folder with no readme of its own would still appear,
with no description, so an undocumented theme stays visible rather
than dropping out of the listing. `_temp` is the exception, and is
absent: it is scratch space rather than a topic category.

The returned table carries an `examples` column as well, too long to
tabulate but useful when deciding where a new dataset belongs.

Then list the layers:

```{r}
list_layers()
```

```
20 layers in 8 themes

 id                                                              title                                         year resolution_m data_type
 biota/grassland_inventory                                       Grassland Inventory for Alberta, Manitoba, a… 2024   30.00      raster   
 biota/natural_regions_subregions_of_alberta                     Natural Regions and Subregions of Alberta     2022      NA      vector   
 boundaries/alberta                                              AB2020_provincial_boundary                    2023      NA      vector   
 elevation/fab_dem                                               FABDEM – Forest And Buildings Removed Copern… 2018  100.00      raster   
 elevation/geomorpho90                                           Alberta Geomorphometric Layers (Geomorpho90m) 2023   90.00      raster   
 elevation/nrcan_mrdem_dsm                                       Medium Resolution Digital Elevation Model (M… 2006   30.00      raster   
 elevation/nrcan_mrdem_dtm                                       Medium Resolution Digital Elevation Model (M… 2006   30.00      raster   
 elevation/nrcan_mrdem_dtm_hillshade                             Medium Resolution Digital Elevation Model (M… 2006   30.00      raster   
 geoscientificInformation/hihydrosoil_v2_ab/abmi1km              HiHydroSoil v2.0                              2020 1000.00      raster   
 geoscientificInformation/soilgrids_250_v2_ab/abmi1km            SoilGrids 2.0 global soil property predictio… 2020 1000.00      raster   
 imageryBaseMapsEarthCover/landsat_summer_mean_indices_2000_2024 Landsat Time Series - Alberta Mean Spectral … 2024   30.00      raster   
 imageryBaseMapsEarthCover/modis_land_cover_dynamics_2001_2023   MODIS Annual Land Cover Dynamics (MCD12Q2) -… 2023  500.00      raster   
 imageryBaseMapsEarthCover/scanfi_v1.2                           SCANFI: Spatialized Canadian National Forest… 2020   30.00      raster   
 inlandWaters/archydro2                                          Alberta ArcHydro Phase 2 Data                 1996  100.00      vector   
 inlandWaters/dynamicSurfaceWaterMaps                            Dynamic Surface Water Maps of Canada from 19… 2023      NA      raster   
 inlandWaters/hydrologically adjusted elevations                 Height Above Nearest Drainage (HAND) - Hydro… 2024   92.77      raster   
 inlandWaters/topographic_wetness_index                          Topographic Wetness Index (TWI)               2024   92.77      raster   
 location/GRID1SQKM_AB2020_gdb                                   GRID1SQKM_AB2020                                NA 1000.00      vector   
 location/GRID1SQKM_AB2020_raster                                GRID1SQKM_AB2020_raster                       2026 1000.00      raster   
 transportation/government_of_alberta_access_layers              Access and Facility Roads - Alberta           2023      NA      vector   
```

Nothing under `_temp` is listed, documented or not — the Sentinel-2
composites staged there are invisible to the catalogue until they are
filed under a theme.

Or narrow the listing to one theme:

```{r}
list_layers(theme = "elevation")
```

```
5 layers in 1 theme

 id                                  title                                         year resolution_m data_type
 elevation/fab_dem                   FABDEM – Forest And Buildings Removed Copern… 2018 100          raster   
 elevation/geomorpho90               Alberta Geomorphometric Layers (Geomorpho90m) 2023  90          raster   
 elevation/nrcan_mrdem_dsm           Medium Resolution Digital Elevation Model (M… 2006  30          raster   
 elevation/nrcan_mrdem_dtm           Medium Resolution Digital Elevation Model (M… 2006  30          raster   
 elevation/nrcan_mrdem_dtm_hillshade Medium Resolution Digital Elevation Model (M… 2006  30          raster   
```

`list_layers()` prints a summary and returns the manifest invisibly.
Assign it, or pass `verbose = FALSE`, to work with the full table —
one row per layer, carrying the parsed template fields alongside
`n_files`, `size_mb`, `data_type`, `path`, and `readme`.

```{r}
cat_df <- list_layers(verbose = FALSE)
dim(cat_df)
#> [1] 18 40
```

Forty columns is too wide to print whole; a few of them give
the shape of the table:

```{r}
cat_df[1:5, c("name", "theme", "year", "resolution_m",
              "n_files", "size_mb", "data_type")]
```

```
                                   name      theme year resolution_m n_files size_mb data_type
1 natural_regions_subregions_of_alberta      biota 2022           NA       1    11.2    vector
2                   grassland_inventory      biota 2024           30       3 30110.4    raster
3                               alberta boundaries 2023           NA       1     0.4    vector
4                               fab_dem  elevation 2018          100       1 11272.9    raster
5                           geomorpho90  elevation 2023           90       1 11197.9    raster
```

The full set of fields — the parsed template alongside the columns
the file scan adds:

```{r}
names(cat_df)
```

```
 [1] "id"                 "name"               "theme"              "product_id"
 [5] "variant"            "title"              "topic_category"     "keywords"
 [9] "year"               "resolution"         "resolution_m"       "xmin"
[13] "xmax"               "ymin"               "ymax"               "crs"
[17] "crs_name"           "datum"              "vertical_crs"       "publication_date"
[21] "start_date"         "end_date"           "format"             "size"
[25] "use_constraints"    "access_constraints" "contact_name"       "contact_email"
[29] "doi"                "online_resource"    "metadata_date"      "abstract"
[33] "purpose"            "credits"            "lineage"            "citation"
[37] "n_files"            "size_mb"            "data_type"          "path"
[41] "readme"             "product_readme"
```

`product_id` and `variant` are what a split record adds. A product
documented in one readme has `variant` `NA` and a `product_id` equal
to its `id`. A product whose data sits in variant subfolders
contributes one row per variant instead of a row of its own: `id`
names the variant folder, `product_id` the product above it, and
`readme` and `product_readme` point at the two halves of the record.

```{r}
cat_df[!is.na(cat_df$variant),
       c("id", "product_id", "variant", "resolution_m", "crs")]
```

```
                                                   id               product_id variant resolution_m       crs
   geoscientificInformation/hihydrosoil_v2_ab/abmi1km  …/hihydrosoil_v2_ab      abmi1km         1000 EPSG:3400
 geoscientificInformation/soilgrids_250_v2_ab/abmi1km  …/soilgrids_250_v2_ab    abmi1km         1000 EPSG:3400
```

The resolution and CRS above come from the variant records; the
titles and licences those rows also carry come from the product
records. `layer_meta()` returns the two read together, so nothing is
lost by the split.

### The catalogue as it stands

The table below is a frozen scan of the share, rendered when this
vignette was built. It is what `list_layers(verbose = FALSE)` returns,
narrowed to the columns worth tabulating — run the call yourself for
the live version and the other thirty-odd columns.

```{r catalogue-table, eval = TRUE, echo = FALSE}
# Written by data-raw/make_catalogue_snapshot.R, already formatted
# for display; missing fields print as empty cells rather than "NA".
snapshot <- readRDS("catalogue_snapshot.rds")

old <- options(knitr.kable.NA = "")
out <- knitr::kable(
  snapshot,
  row.names = FALSE,
  align     = attr(snapshot, "align"),
  caption   = paste0(
    "The spatial data catalogue, scanned ",
    attr(snapshot, "scanned_on"), "."
  )
)
options(old)
out
```

Blank cells are fields the readme leaves unfilled, which is what
keeps a layer out of the matching `find_layer()` filter.

## What variables a product holds

`list_layers()` says which layers exist; `list_variables()` says
what is *inside* them. It reads the band blocks out of each
product readme, so the variables available for extraction can be
browsed without opening a readme or a raster.

```{r}
list_variables(product = "soilgrids_250_v2_ab")
```

```
61 variables across 1 product

 product             band                    measure                                  units         
 soilgrids_250_v2_ab bdod_0-5cm_mean         Bulk density of the fine earth fraction  kg/dm3        
 soilgrids_250_v2_ab bdod_5-15cm_mean        Bulk density of the fine earth fraction  kg/dm3        
 soilgrids_250_v2_ab bdod_15-30cm_mean       Bulk density of the fine earth fraction  kg/dm3        
 soilgrids_250_v2_ab bdod_30-60cm_mean       Bulk density of the fine earth fraction  kg/dm3        
 soilgrids_250_v2_ab bdod_60-100cm_mean      Bulk density of the fine earth fraction  kg/dm3        
 soilgrids_250_v2_ab bdod_100-200cm_mean     Bulk density of the fine earth fraction  kg/dm3        
 soilgrids_250_v2_ab cec_0-5cm_mean          Cation exchange capacity of the soil     cmol(c)/kg    
 soilgrids_250_v2_ab cec_5-15cm_mean         Cation exchange capacity of the soil     cmol(c)/kg    
 ...
```

Bands belong to the product rather than to a variant — resampling
a layer does not change what its bands measure — so a product with
several variants reports one set of rows, not one set per variant.

The compact view above shows two of the columns; the returned
table also carries `scale`, `valid_range`, and `description`.
Called with no arguments it covers the whole catalogue, which
makes it searchable — every band measuring pH, wherever it lives:

```{r}
v <- list_variables(verbose = FALSE)
v[grepl("pH", v$measure, fixed = TRUE),
  c("name", "band", "units")]
```

```
                  name                 band units
71 soilgrids_250_v2_ab     phh2o_0-5cm_mean    pH
72 soilgrids_250_v2_ab    phh2o_5-15cm_mean    pH
73 soilgrids_250_v2_ab   phh2o_15-30cm_mean    pH
74 soilgrids_250_v2_ab   phh2o_30-60cm_mean    pH
75 soilgrids_250_v2_ab  phh2o_60-100cm_mean    pH
76 soilgrids_250_v2_ab phh2o_100-200cm_mean    pH
```

`theme` narrows the listing the same way it does elsewhere. A
product whose readme documents no bands is listed with `band` `NA`
rather than dropped, so the gap stays visible the way
`check_metadata()` keeps missing template fields visible.

## Finding a layer

`find_layer()` searches by what a layer *is* rather than where it
lives. `keyword` matches the folder name, title, keywords, topic
category, and abstract:

```{r}
find_layer(keyword = "elevation")
```

```
10 layers in 4 themes

 id                                                            title                                         year resolution_m data_type
 biota/natural_regions_subregions_of_alberta                   Natural Regions and Subregions of Alberta     2022     NA       vector   
 elevation/fab_dem                                             FABDEM – Forest And Buildings Removed Copern… 2018 100.00       raster   
 elevation/geomorpho90                                         Alberta Geomorphometric Layers (Geomorpho90m) 2023  90.00       raster   
 elevation/nrcan_mrdem_dsm                                     Medium Resolution Digital Elevation Model (M… 2006  30.00       raster   
 elevation/nrcan_mrdem_dtm                                     Medium Resolution Digital Elevation Model (M… 2006  30.00       raster   
 elevation/nrcan_mrdem_dtm_hillshade                           Medium Resolution Digital Elevation Model (M… 2006  30.00       raster   
 imageryBaseMapsEarthCover/modis_land_cover_dynamics_2001_2023 MODIS Annual Land Cover Dynamics (MCD12Q2) -… 2023 500.00       raster   
 inlandWaters/archydro2                                        Alberta ArcHydro Phase 2 Data                 1996 100.00       vector   
 inlandWaters/hydrologically adjusted elevations               Height Above Nearest Drainage (HAND) - Hydro… 2024  92.77       raster   
 inlandWaters/topographic_wetness_index                        Topographic Wetness Index (TWI)               2024  92.77       raster   
```

Note the first hit: a layer filed under `biota` matched on its
abstract, which is the point of searching metadata rather than
folder names.

Filters combine, and each takes either a single value or a
`c(min, max)` range:

```{r}
find_layer(year = c(2020, 2024))
find_layer(resolution = c(0, 120))
find_layer(extent = c(-120, -110, 49, 60))   # decimal degrees
find_layer(theme = "elevation", resolution = 30)
```

A filter narrows to layers *known* to match. Resolution, year, and
extent come from the readme, so a layer whose readme leaves the field
blank is excluded rather than guessed at — `check_metadata()` will
show you which readmes are keeping a layer out.

## Loading a layer

`get_layer()` reads the data with terra, returning a `SpatRaster` or
a `SpatVector` depending on the format:

```{r}
twi <- get_layer("topographic_wetness_index")
nsr <- get_layer("natural_regions_subregions_of_alberta")
```

Pass `return_path = TRUE` when you want the file rather than the
object — to hand it to another tool, or to open it with your own
arguments:

```{r}
get_layer("fab_dem", return_path = TRUE)
#> [1] "//ABMI-DATA2/science/spatial_data/elevation/fab_dem/fab_dem_us_canada.tif"
```

Some folders hold several datasets. `layer_files()` shows what is
there, and `file` picks one:

```{r}
layer_files("grassland_inventory")
#> [1] ".../alberta_grassland_classification_2023.tif"
#> [2] ".../manitoba_grassland_classification_2023.tif"
#> [3] ".../saskatchewan_grassland_classification_2023.tif.tif"

get_layer("grassland_inventory", file = "alberta")
```

Asking for a multi-file layer without `file` is an error that lists
the candidates, so you never silently get the wrong raster. Shapefile
sidecars (`.dbf`, `.shx`, `.prj`) are hidden; pass `all = TRUE` to see
every file in the folder.

Short names work when unambiguous. If two themes hold a folder of the
same name, use the full id:

```{r}
get_layer("elevation/fab_dem")
```

## Reading the metadata

`layer_meta()` prints a layer's provenance, licence, and contact
details, and returns the parsed fields invisibly:

```{r}
layer_meta("fab_dem")
```

```
<sciSpatialR metadata>
  file: //ABMI-DATA2/science/spatial_data/elevation/fab_dem/readme.txt
  Title       FABDEM – Forest And Buildings Removed Copernicus Global DEM…
  Abstract    FABDEM is a global, bare-earth digital elevation model deri…
  Purpose     To provide a globally consistent, near–bare-earth digital e…
  Topic       geoscientificInformation
  Keywords    digital elevation model, DEM, bare-earth, terrain, topograp…
  Resolution  3.23 arc-seconds (~100 m at the equator)
  Published   2022-01-01
  Start       2010-01-01
  End         2018-12-31
  Lineage     FABDEM was derived from the Copernicus GLO-30 Digital Eleva…
  Format      GeoTIFF
  Size        Global dataset; size varies by tile (1° × 1° tiles grouped …
  Use         Creative Commons Attribution–NonCommercial–ShareAlike 4.0 I…
  Access      None. Data is publicly available subject to license terms.
  Contact     Laurence Hawker
  DOI         10.1088/1748-9326/ac4d4f
  Source      https://doi.org/10.1088/1748-9326/ac4d4f
  Updated     2026-01-23
```

Values are truncated for display only; the returned object holds them
in full. Fields the readme left blank are omitted from the printout —
here `Credits` and `Email`, which `check_metadata()` reports below.

Every `Label: value` line in the readme is captured, not just the
template fields, so anything a readme records is reachable. Labels
become `snake_case` keys, and sub-fields indented under a parent are
prefixed with it:

```{r}
md <- layer_meta("fab_dem", print = FALSE)
md$use_constraints
md$extent_west_bounding_coordinate
md$temporal_extent_start_date
```

`read_metadata()` does the same for a readme you name by path, and
`as_metadata_row()` flattens the result to a one-row `data.frame`
with the numeric fields coerced — resolution to metres, bounding
coordinates to decimal degrees, dates to a `year`:

```{r}
md <- read_metadata("path/to/readme.txt")
as_metadata_row(md)
```

Unfilled template placeholders (`[Data Title]`) and non-values
(`"Not Specified"`) are returned as `NA` rather than as text, so a
copied-but-unedited readme reads as missing rather than documented.

## Auditing the metadata

`check_metadata()` scores every layer against the required template
fields and lists the data folders that have no readme at all:

```{r}
check_metadata()
```

```
28 layers audited, 8 with no readme

 id                                                      n_missing complete missing                                      
 …here/climate_na/fab_dem_us_canada_int/Normal_1991_2020 NA        0.000    readme                                       
 …yAtmosphere/climate_na/fab_dem_us_canada_int/Year_2000 NA        0.000    readme                                       
 …yAtmosphere/climate_na/fab_dem_us_canada_int/Year_2005 NA        0.000    readme                                       
 …yAtmosphere/climate_na/fab_dem_us_canada_int/Year_2010 NA        0.000    readme                                       
 …yAtmosphere/climate_na/fab_dem_us_canada_int/Year_2015 NA        0.000    readme                                       
 …yAtmosphere/climate_na/fab_dem_us_canada_int/Year_2020 NA        0.000    readme                                       
 …yAtmosphere/climate_na/fab_dem_us_canada_int/Year_2024 NA        0.000    readme                                       
 inlandWaters/archydro2_merged                           NA        0.000    readme                                       
 biota/grassland_inventory                                8        0.579    purpose, xmin, xmax, ymin, ymax, use_constra…
 boundaries/alberta                                       7        0.632    abstract, credits, resolution, use_constrain…
 location/GRID1SQKM_AB2020_gdb                            7        0.632    abstract, credits, publication_date, use_con…
 imageryBaseMapsEarthCover/scanfi_v1.2                    6        0.684    purpose, xmin, xmax, ymin, ymax, lineage     
 biota/natural_regions_subregions_of_alberta              4        0.789    xmin, xmax, ymin, ymax                       
 elevation/fab_dem                                        2        0.895    credits, contact_email                       
 geoscientificInformation/soilgrids_250_v2_ab/abmi1km     2        0.895    access_constraints, contact_email            
 location/GRID1SQKM_AB2020_raster                         2        0.895    credits, access_constraints                  
 transportation/government_of_alberta_access_layers       2        0.895    resolution, contact_name                     
 elevation/nrcan_mrdem_dsm                                1        0.947    access_constraints                           
 elevation/nrcan_mrdem_dtm                                1        0.947    access_constraints                           
 elevation/nrcan_mrdem_dtm_hillshade                      1        0.947    access_constraints                           
 geoscientificInformation/hihydrosoil_v2_ab/abmi1km       1        0.947    contact_email                                
 inlandWaters/dynamicSurfaceWaterMaps                     1        0.947    resolution                                   
 elevation/geomorpho90                                    0        1.000                                                 
 …seMapsEarthCover/landsat_summer_mean_indices_2000_2024  0        1.000                                                 
 …BaseMapsEarthCover/modis_land_cover_dynamics_2001_2023  0        1.000                                                 
 inlandWaters/archydro2                                   0        1.000                                                 
 inlandWaters/hydrologically adjusted elevations          0        1.000                                                 
 inlandWaters/topographic_wetness_index                   0        1.000                                                 
```

(Abridged; folders with no readme sort first, then the least complete
readmes.) Rows with `missing = "readme"` are folders holding spatial
data that the catalogue cannot see at all — they are the first thing
to fix. Deep ids are trimmed from the front when printed, since the
last segment is what distinguishes one from another; the returned
table holds them in full.

Nothing under `_temp` is audited either, so staging exports are not
counted as undocumented data. The audit stays a list of real gaps.

Use `detail = TRUE` for one row per missing field, which is easier to
tabulate:

```{r}
check_metadata(detail = TRUE)
check_metadata(theme = "elevation", detail = TRUE)
```

```
5 missing fields across 4 layers

 id                                  theme     field             
 elevation/fab_dem                   elevation credits           
 elevation/fab_dem                   elevation contact_email     
 elevation/nrcan_mrdem_dsm           elevation access_constraints
 elevation/nrcan_mrdem_dtm           elevation access_constraints
 elevation/nrcan_mrdem_dtm_hillshade elevation access_constraints
```

Both forms print the compact view above. Pass `verbose = FALSE` for
the plain `data.frame` instead — every column, and the `missing`
list in full rather than truncated to the width of the printout:

```{r}
gaps <- check_metadata(verbose = FALSE)
gaps$missing[gaps$id == "biota/grassland_inventory"]
#> [1] "purpose, xmin, xmax, ymin, ymax, use_constraints, access_constraints, contact_email"
```

## Caching

Scanning a network share is the slow step, so the manifest is built
once per session and reused. After editing a readme or adding data,
rescan:

```{r}
build_catalogue(refresh = TRUE)
```

`build_catalogue(files = FALSE)` skips the file inventory for a
faster, metadata-only scan; `n_files`, `size_mb`, and `data_type` are
then `NA`. Any argument accepted by `build_catalogue()` can be passed
through the query functions:

```{r}
list_layers(root = "D:/spatial_data", refresh = TRUE)
```

## Adding a field to the template

Readmes are parsed generically rather than against a fixed schema,
so a field you add to the template is readable immediately, with no
change to the package. Every `Label: value` line is captured, and a
sub-field indented under a parent label is namespaced by it. Given
this block:

```
Coordinate Reference System:
    Name: NAD83 / Alberta 10-TM (Forest)
    Authority Code: EPSG:3400
    Datum: North American Datum 1983
    Projection: Transverse Mercator
    Vertical CRS: CGVD2013
```

every part is reachable at once:

```{r}
md <- layer_meta("some_layer", print = FALSE)
md$coordinate_reference_system_authority_code
#> [1] "EPSG:3400"
md$coordinate_reference_system_projection
#> [1] "Transverse Mercator"
```

What is *not* automatic is becoming a column of the manifest, and so
a filter in `find_layer()`. That takes one edit: add the parsed key
to `.meta_field_map` in `R/metadata.R`. Adding it to `.meta_required`
in the same file makes `check_metadata()` count it towards
completeness. `R/catalogue.R` builds its columns from that map and
needs no change.

