---
title: "Browsing the spatial data catalogue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Browsing the spatial data catalogue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Chunks are not evaluated: the catalogue reads the ABMI science
# share, which is not reachable when the package is built or
# checked. Output blocks below are transcripts from a real scan.
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

```{r}
library(sciSpatialR)
```

## The share is the catalogue

There is no manifest file to maintain. `build_catalogue()` walks
`\\ABMI-DATA2\science\spatial_data`, parses the readme stored beside
each dataset, and assembles a table from what it finds. Documenting a
dataset means editing its readme; nothing has to be registered
anywhere, and the catalogue cannot drift from the data.

Folders follow the ISO 19115 topic categories and readmes follow the
ABMI spatial metadata template, both documented in the
[geospatial catalog and management guide](https://github.com/bgcasey/geospatial_catalog_and_management_guide).

A folder is catalogued as a **layer** when it holds a readme written
to the dataset template. Theme folders carry a shorter
`Category`/`Description`/`Examples` readme and are reported by
`list_themes()` instead. Datasets sit at whatever depth a theme needs,
so each layer has both a short `name` (`fab_dem`) and a full `id`
(`elevation/fab_dem`).

## Pointing at the share

`spatial_root()` reports where the catalogue reads from and errors if
the share is unreachable.

```{r}
spatial_root()
#> [1] "//ABMI-DATA2/science/spatial_data"
```

To work from a mirror, a mapped drive, or a local copy, set an option
or an environment variable:

```{r}
options(sciSpatialR.spatial_root = "D:/spatial_data")
Sys.setenv(SCISPATIALR_SPATIAL_ROOT = "D:/spatial_data")
```

## Browsing

Start with the themes, which show the topic categories and how much
sits under each:

```{r}
list_themes()
```

```
 theme                     description                                            n_layers
 elevation                 Height above or below sea level.                       5
 inlandWaters              Inland water features, drainage systems, and their …   4
 biota                     Flora and/or fauna in natural environments.            2
 imageryBaseMapsEarthCover Base maps.                                             2
 temp                      <NA>                                                   1
```

A theme with no readme of its own still appears, with no description,
so an undocumented theme folder stays visible rather than dropping
out of the listing.

Then list the layers, optionally within one theme:

```{r}
list_layers()
list_layers(theme = "elevation")
```

`list_layers()` prints a summary and returns the manifest invisibly.
Assign it, or pass `verbose = FALSE`, to work with the full table —
one row per layer, carrying the parsed template fields alongside
`n_files`, `size_mb`, `data_type`, `path`, and `readme`.

```{r}
cat_df <- list_layers(verbose = FALSE)
names(cat_df)
```

## Finding a layer

`find_layer()` searches by what a layer *is* rather than where it
lives. `keyword` matches the folder name, title, keywords, topic
category, and abstract:

```{r}
find_layer(keyword = "elevation")
```

```
10 layers in 4 themes

 id                                                            title                                         year resolution_m data_type
 biota/natural_regions/natural_regions_subregions_of_alberta   Natural Regions and Subregions of Alberta     2022     NA       vector
 elevation/fab_dem                                             FABDEM – Forest And Buildings Removed Copern… 2018 100.00       raster
 elevation/geomorpho90                                         Alberta Geomorphometric Layers (Geomorpho90m) 2023  90.00       raster
 elevation/nrcan_mrdem_dsm                                     Medium Resolution Digital Elevation Model (M… 2006  30.00       raster
 elevation/nrcan_mrdem_dtm                                     Medium Resolution Digital Elevation Model (M… 2006  30.00       raster
 elevation/nrcan_mrdem_dtm_hillshade                           Medium Resolution Digital Elevation Model (M… 2006  30.00       raster
 imageryBaseMapsEarthCover/modis_land_cover_dynamics_2001_2023 MODIS Annual Land Cover Dynamics (MCD12Q2) -… 2023 500.00       raster
 inlandWaters/hydrologically adjusted elevations               Height Above Nearest Drainage (HAND) - Hydro… 2024  92.77       raster
 inlandWaters/streams/archydro2                                Alberta ArcHydro Phase 2 Data                 1996 100.00       vector
 inlandWaters/topographic_wetness_index                        Topographic Wetness Index (TWI)               2024  92.77       raster
```

Note the first hit: a layer filed under `biota` matched on its
abstract, which is the point of searching metadata rather than
folder names.

Filters combine, and each takes either a single value or a
`c(min, max)` range:

```{r}
find_layer(year = c(2020, 2024))
find_layer(resolution = c(0, 120))
find_layer(extent = c(-120, -110, 49, 60))   # decimal degrees
find_layer(theme = "elevation", resolution = 30)
```

A filter narrows to layers *known* to match. Resolution, year, and
extent come from the readme, so a layer whose readme leaves the field
blank is excluded rather than guessed at — `check_metadata()` will
show you which readmes are keeping a layer out.

## Loading a layer

`get_layer()` reads the data with terra, returning a `SpatRaster` or
a `SpatVector` depending on the format:

```{r}
twi <- get_layer("topographic_wetness_index")
nsr <- get_layer("natural_regions_subregions_of_alberta")
```

Pass `return_path = TRUE` when you want the file rather than the
object — to hand it to another tool, or to open it with your own
arguments:

```{r}
get_layer("fab_dem", return_path = TRUE)
#> [1] "//ABMI-DATA2/science/spatial_data/elevation/fab_dem/fab_dem_us_canada.tif"
```

Some folders hold several datasets. `layer_files()` shows what is
there, and `file` picks one:

```{r}
layer_files("grassland_inventory")
#> [1] ".../alberta_grassland_classification_2023.tif"
#> [2] ".../manitoba_grassland_classification_2023.tif"
#> [3] ".../saskatchewan_grassland_classification_2023.tif.tif"

get_layer("grassland_inventory", file = "alberta")
```

Asking for a multi-file layer without `file` is an error that lists
the candidates, so you never silently get the wrong raster. Shapefile
sidecars (`.dbf`, `.shx`, `.prj`) are hidden; pass `all = TRUE` to see
every file in the folder.

Short names work when unambiguous. If two themes hold a folder of the
same name, use the full id:

```{r}
get_layer("elevation/fab_dem")
```

## Reading the metadata

`layer_meta()` prints a layer's provenance, licence, and contact
details, and returns the parsed fields invisibly:

```{r}
layer_meta("fab_dem")
```

```
<sciSpatialR metadata>
  file: //ABMI-DATA2/science/spatial_data/elevation/fab_dem/readme.txt
  Title       FABDEM – Forest And Buildings Removed Copernicus Global DEM…
  Abstract    FABDEM is a global, bare-earth digital elevation model deri…
  Purpose     To provide a globally consistent, near–bare-earth digital e…
  Topic       geoscientificInformation
  Keywords    digital elevation model, DEM, bare-earth, terrain, topograp…
  Resolution  3.23 arc-seconds (~100 m at the equator)
  Published   2022-01-01
  Start       2010-01-01
  End         2018-12-31
  Lineage     FABDEM was derived from the Copernicus GLO-30 Digital Eleva…
  Format      GeoTIFF
  Size        Global dataset; size varies by tile (1° × 1° tiles grouped …
  Use         Creative Commons Attribution–NonCommercial–ShareAlike 4.0 I…
  Access      None. Data is publicly available subject to license terms.
  Contact     Laurence Hawker
  DOI         10.1088/1748-9326/ac4d4f
  Source      https://doi.org/10.1088/1748-9326/ac4d4f
  Updated     2026-01-23
```

Values are truncated for display only; the returned object holds them
in full. Fields the readme left blank are omitted from the printout —
here `Credits` and `Email`, which `check_metadata()` reports below.

Every `Label: value` line in the readme is captured, not just the
template fields, so anything a readme records is reachable. Labels
become `snake_case` keys, and sub-fields indented under a parent are
prefixed with it:

```{r}
md <- layer_meta("fab_dem", print = FALSE)
md$use_constraints
md$extent_west_bounding_coordinate
md$temporal_extent_start_date
```

`read_metadata()` does the same for a readme you name by path, and
`as_metadata_row()` flattens the result to a one-row `data.frame`
with the numeric fields coerced — resolution to metres, bounding
coordinates to decimal degrees, dates to a `year`:

```{r}
md <- read_metadata("path/to/readme.txt")
as_metadata_row(md)
```

Unfilled template placeholders (`[Data Title]`) and non-values
(`"Not Specified"`) are returned as `NA` rather than as text, so a
copied-but-unedited readme reads as missing rather than documented.

## Auditing the metadata

`check_metadata()` scores every layer against the required template
fields and lists the data folders that have no readme at all:

```{r}
check_metadata()
```

```
 id                                    n_missing complete missing
 temp/distance_to_water                NA        0.000    readme
 temp/scanfi_v2/SCANFI_age_v2_20260119 NA        0.000    readme
 biota/vegetation/grassland_inventory   8        0.579    purpose, xmin, xmax, ymin, ymax, use_constraints, …
 elevation/fab_dem                      2        0.895    credits, contact_email
 elevation/geomorpho90                  0        1.000
```

(Abridged; folders with no readme sort first, then the least complete
readmes.) Rows with `missing = "readme"` are folders holding spatial
data that the catalogue cannot see at all — they are the first thing
to fix.
Use `detail = TRUE` for one row per missing field, which is easier to
tabulate:

```{r}
check_metadata(detail = TRUE)
check_metadata(theme = "elevation", detail = TRUE)
```

## Caching

Scanning a network share is the slow step, so the manifest is built
once per session and reused. After editing a readme or adding data,
rescan:

```{r}
build_catalogue(refresh = TRUE)
```

`build_catalogue(files = FALSE)` skips the file inventory for a
faster, metadata-only scan; `n_files`, `size_mb`, and `data_type` are
then `NA`. Any argument accepted by `build_catalogue()` can be passed
through the query functions:

```{r}
list_layers(root = "D:/spatial_data", refresh = TRUE)
```

## Known gap

The metadata template has no CRS element, so `find_layer(crs = )`
matches only the few readmes that happen to record one, and the
`crs` column is otherwise `NA`. Several readmes also leave the
bounding coordinates blank, which excludes them from `extent`
filters. Both would be resolved either by adding the fields to the
template or by reading them from the file headers with `terra::crs()`
and `terra::ext()`.
