---
title: "Extracting covariates at survey locations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting covariates at survey locations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Every chunk here is evaluated. The reference layers ship with the
# package and the demonstration layers are built in the vignette, so
# nothing reaches the network or the ABMI share.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r}
library(sciSpatialR)
library(terra)
library(sf)
```

## Four ways to give a point a value

Harmonization ends with layers that share a grid. Modelling starts
with a table: one row per survey location, one column per
covariate. The extraction functions are the bridge, and which one
you want depends on what the covariate is supposed to mean.

| Question | Call | Returns |
|---|---|---|
| What is the value *at* this location? | `extract_points()` | one column per raster layer |
| What is the average (or sum, or spread) *around* it? | `extract_buffer()` | one column per layer × radius |
| What is the *composition* around it? | `extract_proportion()` | one column per class |
| Which polygon is it *in*? | `extract_vector()` | the polygon's attributes |

The first three read rasters and return a plain `data.frame`; the
last reads a polygon layer and returns an `sf`. The three raster
functions take points as either `sf` or `SpatVector` and reproject
them to the raster when the CRS does not match;
`extract_vector()` takes `sf` on both sides and reprojects the
other way. Either way a point set that has been through
`harmonize_crs()` and a layer that has been through
`resample_to_grid()` need no further preparation.

## Layers and points to extract from

Everything below runs on synthetic layers built on the reference
grid, so this vignette needs no external data. A continuous
surface standing in for a canopy height model, and a four-class
categorical layer derived from it:

```{r}
set.seed(42)

ref   <- ab_grid()
east  <- init(ref, "x")
north <- init(ref, "y")

chm <- 14 +
  6 * sin(east / 70000) * cos(north / 110000) -
  4 * (north - 5.4e6) / 1.2e6
chm <- chm + init(ref, function(n) rnorm(n, 0, 1.5))
chm <- mask(chm, ref)
names(chm) <- "canopy_height"

cover <- classify(
  chm,
  matrix(
    c(-Inf,  8, 1,
         8, 14, 2,
        14, 18, 3,
        18, Inf, 4),
    ncol = 3, byrow = TRUE
  )
)
names(cover) <- "cover"
cover
```

Both are masked to Alberta, which matters later: cells outside the
province are `NA`, and a buffer near the border will contain some
of them.

The survey locations are five sites in geographic coordinates —
the state point data usually arrives in, straight off a GPS or out
of a database:

```{r}
sites <- st_as_sf(
  data.frame(
    site = c("calgary", "edmonton", "fort_mcmurray",
             "grande_prairie", "border"),
    x = c(-114.07, -113.49, -111.38, -118.80, -119.90),
    y = c(  51.05,   53.55,   56.73,    55.17,    54.00)
  ),
  coords = c("x", "y"),
  crs    = 4326
)
sites
```

`border` sits close to the British Columbia line and is there to
show what happens at the edge of a masked layer.

## `extract_points()`

The value of the cell each point falls in:

```{r}
extract_points(chm, sites)
```

Note what did *not* have to happen first. The points are in
EPSG:4326 and the raster is in Alberta 10-TM, so the function
reprojected the points and warned about it. The warning is
deliberate — reprojecting silently is how a point set ends up
somewhere unintended — but it is noise once the reprojection is
known to be right. Harmonize first and the call is quiet:

```{r}
sites_ab <- harmonize_crs(sites)
```

```{r}
extract_points(chm, sites_ab)
```

**The result is a `data.frame`, not an `sf`.** `bind = TRUE`
column-binds the extracted values to the *attributes* of `points`,
and the geometry is dropped along the way. That is usually what a
modelling table wants; when the geometry is needed too, bind the
values back onto the `sf` object yourself:

```{r}
vals      <- extract_points(chm, sites_ab, bind = FALSE)
sites_val <- cbind(sites_ab, vals)
class(sites_val)
```

`bind = FALSE` returns just the extracted columns, in the same row
order as `points`, which is what makes that `cbind` safe.

Multi-layer rasters give one column per layer, so a stack of
prepared covariates is one call:

```{r}
covariates <- c(chm, cover)
extract_points(covariates, sites_ab)
```

Arguments in `...` reach `terra::extract()`. The useful one is
`method`: the default `"simple"` takes the value of the containing
cell, while `"bilinear"` interpolates from the four nearest cell
centres, which is smoother for a continuous surface and meaningless
for class codes.

```{r}
extract_points(chm, sites_ab, method = "bilinear")
```

A point outside the layer's data — off the extent, or on a masked
cell — returns `NA` rather than an error:

```{r}
off <- st_as_sf(
  data.frame(site = "saskatoon", x = -106.67, y = 52.13),
  coords = c("x", "y"), crs = 4326
)
extract_points(chm, harmonize_crs(off, warn = FALSE))
```

That `NA` is worth checking for after every extraction. It means
one of two very different things — the point is outside the study
area, or the layer has a hole in it — and only the second is a
problem with the layer.

## `extract_buffer()`

A point value describes one 1 km cell. Most covariates in a
species model are not that: bird detections respond to the forest
*around* the station, not the pixel under it. `extract_buffer()`
summarises the cells whose centres fall within a circular buffer,
and is vectorised over radii so a whole multi-scale set comes back
in one table:

```{r}
extract_buffer(chm, sites_ab, radii = c(1000, 5000, 20000))
```

Radii are in the CRS units — metres here, because `ab_crs()` is
projected. The columns are named `<layer>_r<radius>_<fun>`, so
several radii, several layers, and several summary functions can
coexist in one table without collision.

Look at the `border` row: the two smaller radii came back fine and
the 20 km one did not. That buffer reaches across the provincial
line into cells that are `NA` in a masked layer, and a mean over
anything containing an `NA` is not a number. Arguments in `...` are
forwarded to the summary function, so `na.rm = TRUE` fixes it:

```{r}
extract_buffer(chm, sites_ab, radii = c(1000, 5000, 20000),
               na.rm = TRUE)
```

**This is the argument to think about rather than to always pass.**
`na.rm = TRUE` means "summarise the part of the buffer I have
data for" — right for a site near the provincial border, where the
missing cells are genuinely outside the study area, and wrong for a
site in the middle of a hole in a layer, where it quietly reports a
20 km mean computed from a handful of cells. The honest version of
that check is to extract a count alongside the summary:

```{r}
extract_buffer(chm, sites_ab, radii = 20000,
               fun = function(v, ...) sum(!is.na(v)),
               fun_name = "n_valid")
```

A 20 km buffer on a 1 km grid covers roughly 1,250 cells; anything
much below that is a buffer that has left the layer.

Two things about that call. `fun` is any function that reduces a
vector to one number, so `sd`, `max`, `median`, or a custom one all
work. And `fun_name` supplies the column suffix: a named function
is labelled from its own name, but an inline function has no name
to take, so pass one rather than letting it fall back to `stat`.

```{r}
extract_buffer(chm, sites_ab, radii = 5000, fun = sd, na.rm = TRUE)
```

Radius is a modelling choice, not a technical one, and the point of
returning several at once is to leave it open. A stack and a set of
radii multiply out:

```{r}
multi <- extract_buffer(covariates, sites_ab,
                        radii = c(5000, 20000), na.rm = TRUE)
names(multi)
```

Note `cover_r5000_mean` in there. It is the mean of the class codes
1–4, which is not a quantity — the mean of "wetland" and "upland"
does not exist. Categorical layers need the next function.

## `extract_proportion()`

For a categorical layer, the meaningful summary within a buffer is
composition: how much of each class. `extract_proportion()` returns
one column per class, and one row per point:

```{r}
extract_proportion(cover, sites_ab, radius = 10000)
```

Each row sums to one:

```{r}
props <- extract_proportion(cover, sites_ab, radius = 10000,
                            bind = FALSE)
rowSums(props)
```

It sums to one for `border` as well, even though 16 of the 298
cells in that buffer are outside the province. Proportions are of
the *valid* cells in the buffer — `NA` cells are dropped before the
tabulation rather than counted as a class — which is the same decision
`na.rm = TRUE` makes in `extract_buffer()`, taken for you because
there is no sensible alternative. The consequence is the same
caveat: a proportion says nothing about how much data it was
computed from. Pair it with a count when a buffer might be mostly
missing.

Radius changes the answer, and a small one concentrates it on the
class under the point — a 2 km buffer on a 1 km grid is about a
dozen cells, so the proportions are coarse fractions of a small
denominator:

```{r}
extract_proportion(cover, sites_ab, radius = 2000)
```

Two constraints follow from how the columns are built, and the
table above has already shown one of them.

**The columns are the classes observed across the buffers, not the
classes in the layer.** `cover_4` is in the 10 km table and gone
from the 2 km one: no 2 km buffer happened to contain that class.
Two runs on different point sets can therefore come back with
different columns, which will misalign silently if you `rbind` them
or feed them to a model fitted on the other set. A single site is
the extreme case:

```{r}
one <- extract_proportion(cover, sites_ab[1, ], radius = 2000,
                          bind = FALSE)
names(one)
names(props)
```

When that matters, reconcile the columns yourself against the full
class set of the layer:

```{r}
classes  <- sort(unique(values(cover, na.rm = TRUE)))
expected <- paste0("cover_", classes)
missing  <- setdiff(expected, names(one))
one[missing] <- 0
one[expected]
```

**The layer must be a single band.** Class proportions for a stack
would need one column set per layer, which the return shape has no
room for, so it errors rather than guessing:

```{r, error = TRUE}
extract_proportion(covariates, sites_ab, radius = 5000)
```

Call it once per categorical layer and `cbind` the results — and
note that `radius` is a single number too, so multiple scales are
multiple calls, unlike `extract_buffer()`.

## `extract_vector()`

The last question is a join rather than a summary: which polygon
does this point fall in? Natural subregion, LUF zone, land
ownership, watershed — anything that arrives as a polygon layer
with attributes.

The demonstration layer is a synthetic set of management zones
covering the province:

```{r}
zones <- st_sf(
  zone_id   = 1:6,
  zone_name = c("southwest", "southeast", "central_west",
                "central_east", "northwest", "northeast"),
  crew      = c("A", "A", "B", "B", "C", "C"),
  geometry  = st_make_grid(st_as_sf(ab_boundary()), n = c(2, 3))
)
zones
```

```{r}
extract_vector(sites_ab, zones)
```

Three differences from the raster extractors are worth naming.

**It is `sf` in, `sf` out.** Both arguments must be `sf` objects —
a `SpatVector` errors here, where the raster functions would have
accepted it — and the geometry survives, because a spatial join is
`sf::st_join()` underneath and the points keep their own geometry
column. That also means the result can be passed straight into
another `extract_vector()` call, or written to a GeoPackage.

**It handles the CRS the other way round.** The raster functions
reproject the points to the raster; this one reprojects the
*polygons* to the points, so the output geometry is always in the
CRS the points arrived in. Passing the unprojected points gives the
same join, in EPSG:4326:

```{r}
st_crs(extract_vector(sites, zones))$input
```

**`cols` trims the join.** A subregion layer with forty attribute
columns will otherwise drag all forty into the modelling table:

```{r}
extract_vector(sites_ab, zones, cols = "zone_name")
```

A name that is not in `polygons` is dropped rather than flagged, so
a typo costs you a column silently — check the result's names, or
`cols` against `names(polygons)`, rather than trusting the call:

```{r}
names(extract_vector(sites_ab, zones, cols = c("zone_name", "zoen")))
```

### Two failure modes

A point outside every polygon joins to `NA`, because the underlying
join is a left join — the row survives, the attributes do not:

```{r}
extract_vector(harmonize_crs(off, warn = FALSE), zones)
```

That is the right default for a modelling table, where a dropped
row is worse than a visible `NA`. `sf::st_join()` arguments reach
through `...`, so `left = FALSE` turns it into an inner join when
you would rather lose the row:

```{r}
nrow(extract_vector(harmonize_crs(off, warn = FALSE), zones,
                    left = FALSE))
```

The second one is the one to watch for: a point inside
*overlapping* polygons is returned once per match, so the result
can be longer than the input:

```{r}
overlapping <- st_sf(
  layer    = c("survey_2019", "survey_2024"),
  geometry = c(
    st_as_sfc(st_bbox(zones)),
    st_as_sfc(st_bbox(zones))
  ),
  crs = st_crs(zones)
)
nrow(sites_ab)
nrow(extract_vector(sites_ab, overlapping))
```

Nested or overlapping polygon layers are common — protected areas
inside land-use zones, historical survey footprints — so check
`nrow()` against the input rather than assuming a one-to-one join.

`st_join()`'s `largest = TRUE` collapses the duplicates back to one
row per point, but read what it does before reaching for it: it
keeps the match with the largest area of overlap, and a point has
*no* area, so every candidate ties and the winner is whichever
polygon comes first in the layer. Here that is the outer zone in one
ordering and the inner zone in the other:

```{r}
nested <- st_sf(
  zone     = c("outer", "inner"),
  geometry = st_sfc(
    st_as_sfc(st_bbox(zones))[[1]],
    st_buffer(st_geometry(sites_ab)[1], 50000)[[1]],
    crs = st_crs(zones)
  )
)

extract_vector(sites_ab[1, ], nested, largest = TRUE)$zone
extract_vector(sites_ab[1, ], nested[2:1, ], largest = TRUE)$zone
```

It is a deduplication, not a resolution. When one of the polygons is
the right answer — the subregion rather than the survey footprint —
subset the polygon layer before the join instead.

## A worked pipeline

The four together, as they would appear at the end of a covariate
script — a point set and a stack of prepared layers going in, one
modelling table coming out:

```{r}
# 1. Points into the reference CRS, once, up front
survey <- harmonize_crs(sites)

# 2. Point-in-cell values for the continuous layers
at_point <- extract_points(chm, survey, bind = FALSE)

# 3. Multi-scale neighbourhood summaries
around <- extract_buffer(chm, survey, radii = c(1000, 5000),
                         na.rm = TRUE, bind = FALSE)

# 4. Composition of the categorical layer at the larger scale
comp <- extract_proportion(cover, survey, radius = 5000,
                           bind = FALSE)

# 5. Which management zone each site is in
in_zone <- extract_vector(survey, zones, cols = "zone_name")

# 6. One table
model_data <- cbind(
  st_drop_geometry(in_zone),
  at_point, around, comp
)
str(model_data)
```

Every step used `bind = FALSE` and the binding happened once at the
end. That is the pattern to prefer: each function returns its
columns in the input row order, so a single `cbind` is safe, whereas
binding at every step carries the site attributes through four
times and leaves you deduplicating columns.

Then confirm the table before it becomes a model:

```{r}
colSums(is.na(model_data))
```

`NA`s here would mean a site outside a layer, a buffer that left the
province, or a point outside every polygon — worth resolving at this
stage rather than discovering as dropped rows in a model fit.

## Cost

Extraction cost scales with the number of cells read, not with the
number of points, which makes the radii the expensive choice.
Doubling a radius quadruples the cells in each buffer, and each
radius is a separate pass over the raster — `radii = c(1000, 5000,
20000)` is three passes, and the 20 km one reads 400 times the
cells of the 1 km one.

For a large point set against province-wide layers, that means:
extract at as few radii as the question needs, prefer one call with
a multi-layer `SpatRaster` over one call per layer, and crop the
raster to the extent of the points first when the points cover a
small part of the province.

## Known gaps

`extract_points()`, `extract_buffer()`, and `extract_proportion()`
drop the point geometry when `bind = TRUE`, returning a
`data.frame` rather than an `sf`. Use `bind = FALSE` and bind the
columns back yourself when the geometry is needed.

`extract_buffer()` includes a cell when its centre falls inside the
buffer, with no partial-cell weighting; a buffer smaller than one
cell falls back to the cells it touches, so the returned value is
that of a handful of whole cells rather than of the circle. Neither
it nor `extract_proportion()` reports how many valid cells went
into a summary — extract a count alongside when that matters.

`extract_proportion()` takes a single layer and a single radius per
call, and derives its class columns from the classes it happens to
see, as shown above.

`extract_vector()` neither collapses duplicate matches nor reports
them, so the row count is yours to check, and it drops `cols` names
that are not in `polygons` without a warning. It also joins by
containment only — a point that misses every polygon by a metre
gets `NA` rather than the nearest one, which for survey coordinates
near a boundary is worth an explicit `sf::st_nearest_feature()`
pass rather than a shrug.

Nothing here handles a temporal dimension: matching a survey to the
covariate layer for its year is left to the calling script, which
typically means one extraction per year and an `rbind`.
