---
title: "Harmonizing layers to the reference grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Harmonizing layers to the reference grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r}
library(sciSpatialR)
library(terra)
library(sf)
```

## One grid, three questions

Covariates arrive from wherever they were published — a global DEM
in geographic coordinates, a national land cover product at 30 m, a
provincial polygon layer. Before any of them can be stacked,
extracted from, or compared, they have to agree on the following properties:
coordinate reference system, extent, resolution, and origin.

The harmonization functions in this package address the following:

1. **What am I aligning to?** The reference layers — `ab_crs()`,
   `ab_boundary()`, `ab_grid()`.
2. **Does this layer already align?** `check_alignment()`.
3. **How do I make it align?** `harmonize_crs()`,
   `resample_to_grid()`, `mask_to_boundary()`.

## The reference layers

### `ab_crs()`

The CRS everything harmonizes to: NAD83 / Alberta 10-TM (Forest).

```{r}
ab_crs()
```

It is a projected CRS in metres, which is what makes the rest of
the package sane — resolutions, buffer radii, and areas are all in
the same unit, and no function has to reason about degrees varying
with latitude.

### `ab_grid()`

The ABMI 1 km reference grid, returned as a template `SpatRaster`:

```{r}
ref <- ab_grid()
ref
```

Cells covering Alberta hold `1`; everything else is `NA`. Of the
857,630 cells in the rectangle, the province occupies:

```{r}
global(ref, "notNA")
```

Two properties are worth knowing before you rely on it.

The template is *snapped to the lattice of the ABMI's 1 km grid
polygons ("\\ABMI-DATA2\science\spatial_data\location\referenceGrids\GRID1SQKM_AB2020_gdb")*, not to a round number. That is why the origin is an odd
pair of values rather than `0, 0`:

```{r}
origin(ref)
```

The payoff is that a cell's raster row and column reproduce its
source `GRID_LABEL`, so raster work and the ABMI grid
tabulations refer to the same cells.

Every raster cell is a full
1 km², whereas the source polygons along the provincial boundary
are clipped to less than that. Area-weighted work on edge cells
should go back to the source polygons on the share. Provenance for
both layers is in
[`inst/extdata/README.md`](https://github.com/bgcasey/sciSpatialR/blob/main/inst/extdata/README.md).

### `ab_boundary()`

The 2020 provincial boundary as one dissolved polygon, in the same
CRS:

```{r}
ab_boundary()
```

It carries no attributes on purpose — it exists to clip, not to
join to.

## Checking alignment

`check_alignment()` compares a layer against the reference on all
four properties and messages the result. The reference grid
agrees with itself:

```{r}
check_alignment(ref)
```

Something freshly downloaded usually does not. Here is a stand-in
for one — a synthetic temperature surface in geographic coordinates:

```{r}
src <- rast(
  nrows = 120, ncols = 90,
  xmin = -121, xmax = -109,
  ymin = 48.5, ymax = 60.5,
  crs  = "EPSG:4326"
)
src <- 25 - 0.55 * init(src, "y") + 2 * sin(init(src, "x") / 2)
names(src) <- "mean_temp"
src
```

```{r}
check_alignment(src)
```

Four mismatches. The function
returns the results invisibly as a named logical vector, so it can
act as TRUE/FALSE gate in a code pipeline.

```{r}
ok <- check_alignment(src, verbose = FALSE)
ok
all(ok)
```

`sf` objects are accepted too, but only the CRS is meaningful for
them. A point layer has no resolution or origin to compare, so
those checks are skipped rather than silently reported as failures:

```{r}
pts <- st_as_sf(
  data.frame(
    site = c("calgary", "edmonton"),
    x    = c(-114.07, -113.49),
    y    = c(51.05, 53.55)
  ),
  coords = c("x", "y"),
  crs    = 4326
)
check_alignment(pts)
```

## Get the CRS right first

**This is the step to not skip.** `resample_to_grid()` wraps
`terra::resample()`, which aligns a layer to a template but does
*not* reproject. Handed a layer in the wrong CRS, it treats the
reference extent as though it were in the layer's own coordinates.

```{r}
src_ab <- project(src, ab_crs())      # bilinear, fine for continuous
check_alignment(src_ab, verbose = FALSE)[["crs"]]
```

The CRS now agrees; the grid geometry still does not, which is what
`resample_to_grid()` is for.

## Harmonizing rasters

### `resample_to_grid()`

With the CRS harmonized, resampling snaps the layer onto the reference
lattice:

```{r}
temp_1km <- resample_to_grid(src_ab)
temp_1km
aligned <- check_alignment(temp_1km, verbose = FALSE)
aligned
```

All four properties now agree, which is the whole point: any two
layers put through this call can be stacked, differenced, or
extracted from together.

Notice the message. `terra::resample()` always returns a layer on
the reference geometry — that part is not in question. What varies
is *how each output cell gets its value*, and the right answer
depends on which way the resolution changes:

| Direction | Continuous | Categorical |
|---|---|---|
| Coarsening (fine to coarse) | `"average"` | `"mode"` |
| Same resolution | `"near"` | `"near"` |
| Refining (coarse to fine) | `"bilinear"` | `"near"` |

`resample_to_grid()` reads `res(x)` against `res(ref)`, checks
whether the layer is categorical, picks from that table, and says
what it picked. Pass `method` to override it at any point.

Coarsening is the case that goes wrong quietly. A 30 m layer has
roughly 1,111 cells inside every 1 km cell: `"average"` reads all
of them, `"bilinear"` reads four, `"near"` reads one. All three
produce a correctly aligned raster, and `check_alignment()` passes
on all three, because alignment is about geometry and this is a
question about values.

```{r}
fine <- rast(
  xmin = 400000, xmax = 430000,
  ymin = 5800000, ymax = 5830000,
  res  = 30, crs = ab_crs()
)
set.seed(1)
values(fine) <- runif(ncell(fine), 0, 30)
win <- crop(ref, ext(400000, 430000, 5800000, 5830000))

by_mean <- resample_to_grid(fine, win)
by_near <- resample_to_grid(fine, win, method = "near")

# both align; they do not agree
all(check_alignment(by_near, win, verbose = FALSE))
mean(abs(values(by_mean) - values(by_near)), na.rm = TRUE)
sd(values(by_mean), na.rm = TRUE)
```

The disagreement is larger than the spread of the layer itself,
which is what picking one cell out of 1,111 costs.

At equal resolution there is nothing to summarise — every output
cell has exactly one input cell, just offset — so `"near"` carries
the values across intact. Interpolating there is pure loss:

```{r}
off_lattice <- shift(win, dx = 300, dy = 300)
values(off_lattice) <- runif(ncell(off_lattice), 0, 30)

same_near <- resample_to_grid(off_lattice, win)
same_bil  <- resample_to_grid(off_lattice, win, method = "bilinear")

sd(values(off_lattice), na.rm = TRUE)   # input
sd(values(same_near),   na.rm = TRUE)   # preserved
sd(values(same_bil),    na.rm = TRUE)   # smoothed away
```

Class codes are the other half of the table. Interpolating between
land cover classes 1 and 3 produces a 2, which is meaningless.
Build a categorical layer from the same surface to see it:

```{r}
zones <- classify(
  src,
  matrix(
    c(-Inf, -8, 1,
        -8, -5, 2,
        -5, -2, 3,
        -2, Inf, 4),
    ncol = 3, byrow = TRUE
  )
)
names(zones) <- "zone"
zones_ab <- project(zones, ab_crs(), method = "near")
sort(unique(values(zones_ab, na.rm = TRUE)))
```

A layer built this way holds class codes as plain numbers, so
nothing marks it as categorical and the automatic choice treats it
as a measurement. The codes become a continuum:

```{r}
wrong <- resample_to_grid(zones_ab, method = "bilinear")
length(unique(values(wrong, na.rm = TRUE)))
range(values(wrong, na.rm = TRUE))
```

Naming the method keeps them intact:

```{r}
zone_1km <- resample_to_grid(zones_ab, method = "near")
sort(unique(values(zone_1km, na.rm = TRUE)))
```

Detection works off `terra::is.factor()`, so a layer with a proper
level table is handled without being told:

```{r}
zones_fct <- zones_ab
levels(zones_fct) <- data.frame(
  value = 1:4,
  zone  = c("cold", "cool", "mild", "warm")
)
auto_zone <- resample_to_grid(zones_fct)
sort(unique(values(auto_zone, na.rm = TRUE)))
```

Anything holding class codes as bare numbers needs `method =`
given explicitly. That is the one thing the automatic choice
cannot see.

### Any resolution ratio

Summarising the fine cells does not require the resolutions to
divide evenly. `terra::resample()` works per output cell — each
1 km cell reads whatever falls inside it, 1,111 cells or 1,123 —
rather than tiling the input into fixed blocks. A 300 m layer,
which does not divide into 1 km at all, lands on the grid exactly:

```{r}
odd <- rast(
  xmin = 400000, xmax = 430300,
  ymin = 5800000, ymax = 5830300,
  res  = 300, crs = ab_crs()
)
odd <- init(odd, "y")
res(odd)

odd_1km <- resample_to_grid(odd, win)
res(odd_1km)
all(check_alignment(odd_1km, win, verbose = FALSE))
```

That is worth stating plainly because the obvious alternative —
`terra::aggregate()` — cannot do it. Aggregation coarsens by an
integer factor, and a factor is a single number: 1000 / 300 rounds
to 3, giving 900 m cells. It also lays its blocks out from the
input's own corner, so it changes cell size without changing
position. Alignment and summary are one operation here, not two.

### Summaries beyond the mean

`method` accepts everything `terra::resample()` does, which covers
most of what a coarse cell might need to say about the fine cells
inside it:

```{r}
totals <- resample_to_grid(fine, win, method = "sum")
spread <- c(
  resample_to_grid(fine, win, method = "min"),
  resample_to_grid(fine, win, method = "max")
)
names(spread) <- c("min", "max")
global(spread, "mean", na.rm = TRUE)
```

`"sum"`, `"min"`, `"q1"`, `"median"`, `"q3"`, `"max"`, and `"rms"`
are all available, alongside `"average"` and `"mode"`.

For a summary not on that list — a standard deviation, an
interquartile range, a function of your own — drop to
`terra::aggregate()`, which takes an arbitrary function, then
resample its output to snap it into place:

```{r}
het <- aggregate(fine, fact = 33, fun = sd)   # 990 m, off-lattice
het_1km <- resample_to_grid(het, win)
all(check_alignment(het_1km, win, verbose = FALSE))
```

Two steps, and only because `sd` is not one of the built-in
methods. For a mean, a mode, or a sum, one call does both jobs.

### Reading the message

`resample_to_grid()` reports its choice and the reason on every
call. In a loop over a folder of covariates that log is the record
of what happened to each layer, and it is where a surprise shows
up — a layer you thought was 1 km turning out to be 250 m, or a
class raster being treated as continuous because its codes are
stored as plain numbers. Pass `quiet = TRUE` to silence it.

The choice is a default, not a constraint. `method` overrides it
whenever the layer's meaning calls for something else — `"sum"`
for counts, `"near"` for a layer whose exact values must survive,
`"bilinear"` for a coarse smooth surface being refined.

### `mask_to_boundary()`

Harmonizing to the grid gives a layer the right geometry but not
the right footprint: a reprojected national product still carries
values across almost the whole reference rectangle, including well
outside Alberta.

```{r}
global(temp_1km, "notNA")
```

Masking clips it to the province. The default boundary is
`"alberta"`, so the common call takes no second argument:

```{r}
temp_ab <- mask_to_boundary(temp_1km)
global(temp_ab, "notNA")
```

`inverse = TRUE` keeps the outside instead — useful for checking
what a clip removed, or for building an outside-Alberta mask:

```{r}
global(mask_to_boundary(temp_1km, inverse = TRUE), "notNA")
```

Any `SpatVector` or `sf` polygon can be passed directly, and it is
reprojected to the raster's CRS when needed, so a study-area polygon
in whatever CRS it arrived in works without preparation:

```{r}
study_area <- vect(
  ext(400000, 500000, 5800000, 5900000),
  crs = ab_crs()
)
global(mask_to_boundary(temp_1km, study_area), "notNA")

study_ll <- project(study_area, "EPSG:4326")
global(mask_to_boundary(temp_1km, study_ll), "notNA")
```

Same count from both: the reprojection happens internally rather
than being left to the caller. Unknown shortcuts fail loudly and
list what is valid:

```{r, error = TRUE}
mask_to_boundary(temp_1km, "saskatchewan")
```

`"natural_regions"` is recognised as a shortcut but the layer is not
yet packaged, so it errors with that explanation and asks for a
polygon instead. Until it ships, pull the natural regions layer from
the catalogue — `get_layer("natural_regions_subregions_of_alberta")`
— and pass it directly.

## Harmonizing points

Points get their own function, because for point data the right
answer is almost always to move the points rather than the raster.
Reprojecting a raster resamples it and introduces artefacts;
reprojecting points is exact.

```{r}
pts_ab <- harmonize_crs(pts)
st_crs(pts_ab)$input
st_coordinates(pts_ab)
```

Note the message. The reporting distinguishes two cases, and the
distinction is deliberate:

- Left at the default — the package reference grid — reprojection
  is the intended outcome, so it is reported as a **message**.
- Given a raster the caller supplied, a CRS mismatch may well be a
  mistake, and reprojecting the raster might have been the better
  choice, so it **warns**.

```{r}
utm <- rast(
  nrows = 10, ncols = 10, crs = "EPSG:32612",
  xmin = 300000, xmax = 400000,
  ymin = 5600000, ymax = 5700000
)
pts_utm <- harmonize_crs(pts, utm)
st_crs(pts_utm)$input
```

Points already in the target CRS are returned untouched, with no
message and no needless transformation, so the call is safe to leave
in a pipeline that runs on both raw and prepared inputs:

```{r}
identical(harmonize_crs(pts_ab), pts_ab)
```

Pass `warn = FALSE` to silence both, once the reprojection is a
known, intended step.

## A worked pipeline

Everything above, as it would appear in a covariate-preparation
script — a downloaded continuous layer and a point set, both brought
onto the reference grid:

```{r}
# 1. Check what arrived
check_alignment(src)

# 2. CRS first; nothing else works until this passes
covariate <- project(src, ab_crs())

# 3. Snap to the reference grid
covariate <- resample_to_grid(covariate)

# 4. Clip to Alberta
covariate <- mask_to_boundary(covariate)

# 5. Confirm, do not assume
check_alignment(covariate)

# 6. Bring the points into the same CRS
sites <- harmonize_crs(pts)
```

```{r}
covariate
```

From here `extract_points()` and the other extraction functions take
over; they are covered separately.

## Working off the default grid

Every reference default is an ordinary argument, so a project on a
different grid — a 250 m analysis, a single natural subregion, work
outside Alberta — overrides it in place:

```{r}
own_ref <- rast(
  xmin = 400000, xmax = 500000,
  ymin = 5800000, ymax = 5900000,
  res  = 250, crs = ab_crs()
)
own <- resample_to_grid(src_ab, ref = own_ref)
all(check_alignment(own, ref = own_ref, verbose = FALSE))
```

`ref` is the argument for `check_alignment()` and
`resample_to_grid()`; `raster` for `harmonize_crs()`; `boundary`
for `mask_to_boundary()`. The
package's own reference layers are just the defaults those arguments
carry, not a requirement — but note that the guarantee at the top of
this vignette weakens accordingly. Layers harmonized to different
references align with each other only if you have arranged it.

## Known gaps

`resample_to_grid()` does not reproject, and fails silently when
handed a mismatched CRS — an all-`NA` raster rather than an error,
as shown above. Calling `check_alignment()` first is currently the
defence; the function could reasonably reproject on the fly, or at
least refuse.

The automatic method choice reads `terra::is.factor()`, so a
categorical layer whose codes are stored as plain numbers is
treated as continuous. Nothing in the raster distinguishes the
two, so this cannot be fixed by inspection — pass `method` for
those layers.

`"natural_regions"` is a recognised `mask_to_boundary()` shortcut
with no packaged layer behind it yet.
