---
title: "Harmonizing layers to the reference grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Harmonizing layers to the reference grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Unlike the catalogue vignette, every chunk here is evaluated: the
# reference layers ship with the package and the demonstration
# layers are built in the vignette, so nothing reaches the network.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r}
library(sciSpatialR)
library(terra)
library(sf)
```

## One grid, three questions

Covariates arrive from wherever they were published — a global DEM
in geographic coordinates, a national land cover product at 30 m, a
provincial polygon layer. Before any of them can be stacked,
extracted from, or compared, they have to agree on four properties:
coordinate reference system, extent, resolution, and origin.

This package answers that with a single default target. Every
harmonization function takes `ref = ab_grid()` and every masking
call defaults to Alberta, so the common case needs no arguments and
layers processed with the package align by construction rather than
by convention.

The workflow is three questions, in order:

1. **What am I aligning to?** The reference layers — `ab_crs()`,
   `ab_boundary()`, `ab_grid()`.
2. **Does this layer already align?** `check_alignment()`.
3. **How do I make it align?** `harmonize_crs()`,
   `resample_to_grid()`, `aggregate_to_grid()`,
   `mask_to_boundary()`.

## The reference layers

### `ab_crs()`

The CRS everything harmonizes to: NAD83 / Alberta 10-TM (Forest).

```{r}
ab_crs()
```

It is a projected CRS in metres, which is what makes the rest of
the package sane — resolutions, buffer radii, and areas are all in
the same unit, and no function has to reason about degrees varying
with latitude.

### `ab_grid()`

The ABMI 1 km reference grid, returned as a template `SpatRaster`:

```{r}
ref <- ab_grid()
ref
```

Cells covering Alberta hold `1`; everything else is `NA`. Of the
857,630 cells in the rectangle, the province occupies:

```{r}
global(ref, "notNA")
```

Two properties are worth knowing before you rely on it.

The template is *snapped to the lattice of the source grid
polygons*, not to a round number. That is why the origin is an odd
pair of values rather than `0, 0`:

```{r}
origin(ref)
```

The payoff is that a cell's raster row and column reproduce its
source `GRID_LABEL` (`"<row>_<col>"`, row 1 at the north edge,
column 1 at the west edge), so raster work and the ABMI grid
tabulations refer to the same cells. Snapping the template to a
tidier origin would break that correspondence.

The second property is a caveat: every raster cell is a full
1 km², whereas the source polygons along the provincial boundary
are clipped to less than that. Area-weighted work on edge cells
should go back to the source polygons on the share. Provenance for
both layers is in
[`inst/extdata/README.md`](https://github.com/bgcasey/sciSpatialR/blob/main/inst/extdata/README.md).

### `ab_boundary()`

The 2020 provincial boundary as one dissolved polygon, in the same
CRS:

```{r}
ab_boundary()
```

It carries no attributes on purpose — it exists to clip, not to
join to. Layers are read from disk on each call rather than cached,
because terra objects hold external pointers that do not survive a
session; both files are small enough that re-reading is cheap.

## Checking alignment

`check_alignment()` compares a layer against the reference on all
four properties and messages the result. The reference grid trivially
agrees with itself:

```{r}
check_alignment(ref)
```

Something freshly downloaded usually does not. Here is a stand-in
for one — a synthetic temperature surface in geographic coordinates,
built so this vignette needs no external data:

```{r}
src <- rast(
  nrows = 120, ncols = 90,
  xmin = -121, xmax = -109,
  ymin = 48.5, ymax = 60.5,
  crs  = "EPSG:4326"
)
src <- 25 - 0.55 * init(src, "y") + 2 * sin(init(src, "x") / 2)
names(src) <- "mean_temp"
src
```

```{r}
check_alignment(src)
```

Four mismatches, which is the normal starting point. The function
returns the results invisibly as a named logical vector, so it can
gate a pipeline as well as inform a human:

```{r}
ok <- check_alignment(src, verbose = FALSE)
ok
all(ok)
```

`sf` objects are accepted too, but only the CRS is meaningful for
them — a point layer has no resolution or origin to compare, and
those checks are skipped rather than silently reported as failures:

```{r}
pts <- st_as_sf(
  data.frame(
    site = c("calgary", "edmonton"),
    x    = c(-114.07, -113.49),
    y    = c(51.05, 53.55)
  ),
  coords = c("x", "y"),
  crs    = 4326
)
check_alignment(pts)
```

Extent, resolution, and origin are compared with a tolerance
(`tol`, default `1e-6`) rather than exactly, since coordinates that
have been through a projection rarely round-trip bit-for-bit.

## Get the CRS right first

**This is the step to not skip.** `resample_to_grid()` wraps
`terra::resample()`, which aligns a layer to a template but does
*not* reproject. Handed a layer in the wrong CRS, it treats the
reference extent as though it were in the layer's own coordinates —
where nothing overlaps — and returns a raster of the right shape
holding nothing at all:

```{r}
bad <- resample_to_grid(src)          # src is still EPSG:4326
global(bad, "notNA")
crs(bad, describe = TRUE)$name
```

No error and no warning: 857,630 cells, none of them data, still
carrying the source CRS. Running `check_alignment()` first is what
catches this, and reprojecting is what fixes it. For rasters that
means `terra::project()`, choosing the method to suit the
measurement:

```{r}
src_ab <- project(src, ab_crs())      # bilinear, fine for continuous
check_alignment(src_ab, verbose = FALSE)[["crs"]]
```

The CRS now agrees; the grid geometry still does not, which is what
`resample_to_grid()` is for.

## Harmonizing rasters

### `resample_to_grid()`

With the CRS settled, resampling snaps the layer onto the reference
lattice:

```{r}
temp_1km <- resample_to_grid(src_ab)
temp_1km
aligned <- check_alignment(temp_1km, verbose = FALSE)
aligned
```

All four properties now agree, which is the whole point: any two
layers put through this call can be stacked, differenced, or
extracted from together without further thought.

The default method is `"bilinear"`, which is right for continuous
measurements and wrong for class codes. Interpolating between land
cover classes 1 and 3 produces a 2 that means nothing. Build a
categorical layer from the same surface to see it:

```{r}
zones <- classify(
  src,
  matrix(
    c(-Inf, -8, 1,
        -8, -5, 2,
        -5, -2, 3,
        -2, Inf, 4),
    ncol = 3, byrow = TRUE
  )
)
names(zones) <- "zone"
zones_ab <- project(zones, ab_crs(), method = "near")
sort(unique(values(zones_ab, na.rm = TRUE)))
```

Resampled with the default method, those four codes become a
continuum:

```{r}
wrong <- resample_to_grid(zones_ab)
length(unique(values(wrong, na.rm = TRUE)))
range(values(wrong, na.rm = TRUE))
```

`categorical = TRUE` forces nearest-neighbour and keeps the codes
intact:

```{r}
zone_1km <- resample_to_grid(zones_ab, categorical = TRUE)
sort(unique(values(zone_1km, na.rm = TRUE)))
```

The flag is a guardrail, not new behaviour — it is equivalent to
`method = "near"`, but it says *why* nearest neighbour was chosen,
which survives review better than a bare method string.

### `aggregate_to_grid()`

Resampling a 30 m layer straight to 1 km samples one fine cell per
coarse cell and discards the other roughly 1,100. When the coarse
value should summarise the fine cells rather than sample them —
mean canopy height, total road length, dominant cover class —
aggregate instead.

`aggregate_to_grid()` derives an integer factor from the resolution
ratio and calls `terra::aggregate()`. Take a window of the reference
grid and a 250 m layer nested inside it:

```{r}
win  <- crop(ref, ext(400000, 500000, 5800000, 5900000))
fine <- disagg(win, fact = 4)         # 250 m, nested in win
fine <- init(fine, "y")               # a surface to summarise
res(fine)
```

```{r}
coarse <- aggregate_to_grid(fine, win)
res(coarse)
check_alignment(coarse, win, verbose = FALSE)[["origin"]]
```

`fun` chooses the summary — `"mean"` by default, or `"sum"`,
`"max"`, `"min"`, or any function `terra::aggregate()` accepts. For
class codes, `categorical = TRUE` forces the modal class, since a
mean of class codes is meaningless in the same way a bilinear
interpolation of them is:

```{r}
fine_cat  <- (init(fine, "row") - 1) %/% 100 + 1   # four bands
modal_1km <- aggregate_to_grid(fine_cat, win, categorical = TRUE)
sort(unique(values(modal_1km, na.rm = TRUE)))
```

Two limitations follow from aggregation working by factor.

**The ratio has to be an integer.** A 300 m layer does not divide
into 1 km, so the factor is rounded and the result lands at 900 m
with a warning rather than silently pretending to be 1 km:

```{r}
odd <- rast(
  xmin = 400000, xmax = 500200,
  ymin = 5800000, ymax = 5900200,
  res  = 300, crs = ab_crs()
)
odd <- init(odd, "y")
res(odd)
agg_odd <- aggregate_to_grid(odd, win)
res(agg_odd)
```

**Aggregation preserves the input's origin.** The factor fixes
resolution, not position, so a layer offset from the reference
lattice aggregates to the right cell size in the wrong place:

```{r}
shifted     <- shift(fine, dx = 125, dy = 125)
agg_shifted <- aggregate_to_grid(shifted, win)

off <- check_alignment(agg_shifted, win, verbose = FALSE)
off
```

Resolution and CRS pass; extent and origin do not. The fix is to
aggregate for the summary statistic, then resample to snap:

```{r}
snapped <- resample_to_grid(agg_shifted, win)
all(check_alignment(snapped, win, verbose = FALSE))
```

Aggregate-then-resample is the general recipe for a fine layer that
is not already nested in the reference lattice: aggregation does the
statistically meaningful part, resampling does the alignment. Always
confirm with `check_alignment()` rather than assuming.

### Which one to use

| Situation | Call |
|---|---|
| Layer coarser than, or near, 1 km | `resample_to_grid()` |
| Layer much finer, and the coarse value should summarise the fine cells | `aggregate_to_grid()`, then `resample_to_grid()` if it does not align |
| Class codes, either way | add `categorical = TRUE` |
| Wrong CRS | `terra::project()` first — neither function reprojects |

### `mask_to_boundary()`

Harmonizing to the grid gives a layer the right geometry but not
the right footprint: a reprojected national product still carries
values across almost the whole reference rectangle, including well
outside Alberta.

```{r}
global(temp_1km, "notNA")
```

Masking clips it to the province. The default boundary is
`"alberta"`, so the common call takes no second argument:

```{r}
temp_ab <- mask_to_boundary(temp_1km)
global(temp_ab, "notNA")
```

`inverse = TRUE` keeps the outside instead — useful for checking
what a clip removed, or for building an outside-Alberta mask:

```{r}
global(mask_to_boundary(temp_1km, inverse = TRUE), "notNA")
```

Any `SpatVector` or `sf` polygon can be passed directly, and it is
reprojected to the raster's CRS when needed, so a study-area polygon
in whatever CRS it arrived in works without preparation:

```{r}
study_area <- vect(
  ext(400000, 500000, 5800000, 5900000),
  crs = ab_crs()
)
global(mask_to_boundary(temp_1km, study_area), "notNA")

study_ll <- project(study_area, "EPSG:4326")
global(mask_to_boundary(temp_1km, study_ll), "notNA")
```

Same count from both: the reprojection happens internally rather
than being left to the caller. Unknown shortcuts fail loudly and
list what is valid:

```{r, error = TRUE}
mask_to_boundary(temp_1km, "saskatchewan")
```

`"natural_regions"` is recognised as a shortcut but the layer is not
yet packaged, so it errors with that explanation and asks for a
polygon instead. Until it ships, pull the natural regions layer from
the catalogue — `get_layer("natural_regions_subregions_of_alberta")`
— and pass it directly.

## Harmonizing points

Points get their own function, because for point data the right
answer is almost always to move the points rather than the raster.
Reprojecting a raster resamples it and introduces artefacts;
reprojecting points is exact.

```{r}
pts_ab <- harmonize_crs(pts)
st_crs(pts_ab)$input
st_coordinates(pts_ab)
```

Note the message. The reporting distinguishes two cases, and the
distinction is deliberate:

- Left at the default — the package reference grid — reprojection
  is the intended outcome, so it is reported as a **message**.
- Given a raster the caller supplied, a CRS mismatch may well be a
  mistake, and reprojecting the raster might have been the better
  choice, so it **warns**.

```{r}
utm <- rast(
  nrows = 10, ncols = 10, crs = "EPSG:32612",
  xmin = 300000, xmax = 400000,
  ymin = 5600000, ymax = 5700000
)
pts_utm <- harmonize_crs(pts, utm)
st_crs(pts_utm)$input
```

Points already in the target CRS are returned untouched, with no
message and no needless transformation, so the call is safe to leave
in a pipeline that runs on both raw and prepared inputs:

```{r}
identical(harmonize_crs(pts_ab), pts_ab)
```

Pass `warn = FALSE` to silence both, once the reprojection is a
known, intended step.

## A worked pipeline

Everything above, as it would appear in a covariate-preparation
script — a downloaded continuous layer and a point set, both brought
onto the reference grid:

```{r}
# 1. Check what arrived
check_alignment(src)

# 2. CRS first; nothing else works until this passes
covariate <- project(src, ab_crs())

# 3. Snap to the reference grid
covariate <- resample_to_grid(covariate)

# 4. Clip to Alberta
covariate <- mask_to_boundary(covariate)

# 5. Confirm, do not assume
check_alignment(covariate)

# 6. Bring the points into the same CRS
sites <- harmonize_crs(pts)
```

```{r}
covariate
```

From here `extract_points()` and the other extraction functions take
over; they are covered separately.

## Working off the default grid

Every reference default is an ordinary argument, so a project on a
different grid — a 250 m analysis, a single natural subregion, work
outside Alberta — overrides it in place:

```{r}
own_ref <- rast(
  xmin = 400000, xmax = 500000,
  ymin = 5800000, ymax = 5900000,
  res  = 250, crs = ab_crs()
)
own <- resample_to_grid(src_ab, ref = own_ref)
all(check_alignment(own, ref = own_ref, verbose = FALSE))
```

`ref` is the argument for `check_alignment()`,
`resample_to_grid()`, and `aggregate_to_grid()`; `raster` for
`harmonize_crs()`; `boundary` for `mask_to_boundary()`. The
package's own reference layers are just the defaults those arguments
carry, not a requirement — but note that the guarantee at the top of
this vignette weakens accordingly. Layers harmonized to different
references align with each other only if you have arranged it.

## Known gaps

`resample_to_grid()` and `aggregate_to_grid()` do not reproject, and
`resample_to_grid()` fails silently when handed a mismatched CRS —
an all-`NA` raster rather than an error, as shown above. Calling
`check_alignment()` first is currently the defence; the functions
could reasonably reproject on the fly, or at least refuse.

`"natural_regions"` is a recognised `mask_to_boundary()` shortcut
with no packaged layer behind it yet.
