---
title: "Inspecting layers before you use them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inspecting layers before you use them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Every chunk here is evaluated. The reference layers ship with the
# package and the demonstration layer is built in the vignette, so
# nothing reaches the network or the ABMI share.
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  dpi        = 96
)
```

```{r}
library(sciSpatialR)
library(terra)
```

## Three questions about a layer

A covariate that has been harmonized still has to be looked at
before it is trusted. Three questions cover most of what goes
wrong, and one function answers each:

1. **What is actually in it?** `raster_stats()` — cell counts,
   missingness, and value summaries as a table.
2. **How are the values distributed?** `plot_hist()` — where they
   pile up, and how long the tails are.
3. **Does the map look right?** `plot_raster()` — the spatial
   pattern, and whether the layer is clipped where it should be.

The first is a table you can assert on in a script; the other two
are `ggplot` objects. All three accept a `SpatRaster` or a path, so
they work on a layer in memory and on a freshly exported GeoTIFF
alike.

## A layer to inspect

Everything below runs on a synthetic surface on the reference grid,
built so this vignette needs no external data. It is deliberately
imperfect: a handful of absurd outliers, and a rectangular hole
where an upstream process dropped cells.

```{r}
set.seed(42)

ref   <- ab_grid()
east  <- init(ref, "x")
north <- init(ref, "y")

# A smooth surface with some structure to look at
chm <- 14 +
  6 * sin(east / 70000) * cos(north / 110000) -
  4 * (north - 5.4e6) / 1.2e6
chm <- chm + init(ref, function(n) rnorm(n, 0, 0.6))

# Sensor glitches: a few cells with impossible values
glitch <- sample(which(!is.na(values(ref, mat = FALSE))), 40)
chm[glitch] <- 250

# A gap left by an upstream process
chm[300:360, 180:240] <- NA

chm <- mask(chm, ref)
names(chm) <- "canopy_height"
chm
```

## `raster_stats()`

The table first, because it is the cheapest and the most
assertable. Counts, minimum, maximum, mean, and standard deviation
come from `terra::global()`, which streams the raster from disk
rather than loading it, so this scales to province-wide layers.

```{r}
raster_stats(chm)
```

Every column is defined above the table. That printing is the
`verbose` argument, on by default; turn it off when the table is
going somewhere other than a human:

```{r}
stats <- raster_stats(chm, verbose = FALSE)
stats$pct_na
```

Two things in that row are worth a second look. `pct_na` is high
because the reference grid rectangle is mostly outside Alberta —
expected here, but the same number on a layer that should cover the
province is the first sign of a bad clip. And `max` is 250 against
a mean under 12, which is the glitch showing up.

Quantiles make the tail explicit. They are optional because, unlike
the other statistics, they need the values themselves:

```{r}
raster_stats(chm, quantiles = c(0.02, 0.5, 0.98), verbose = FALSE)
```

The 98th percentile sits far below the maximum, so the extremes are
a handful of cells rather than a real upper tail. Those columns are
computed from a regular sample when a layer has more than `maxcell`
cells, so they are approximate on large rasters; the counts and the
min/max/mean/sd never are.

### A folder of exports

`x` can also be a path, or a directory, which is the point of the
`source` column: a QA pass over everything that came back from a
batch export is one call.

```{r}
export_dir <- file.path(tempdir(), "exports")
dir.create(export_dir, showWarnings = FALSE)

writeRaster(
  aggregate(chm, 8, na.rm = TRUE),
  file.path(export_dir, "canopy_height_8km.tif"),
  overwrite = TRUE
)
writeRaster(
  aggregate(chm, 16, fun = "max", na.rm = TRUE),
  file.path(export_dir, "canopy_max_16km.tif"),
  overwrite = TRUE
)

raster_stats(export_dir, verbose = FALSE)
```

Both rows carry the same layer name, because both files inherited
the band name from the raster they were written from; `source` is
what tells them apart. It is `NA` for a raster held in memory. A
band written *without* a name would come back as terra's `lyr.1`,
so those are labelled with the file name instead.

An empty layer returns `NA` for the value summaries rather than the
`NaN` and `Inf` that `min()` and `mean()` produce on nothing:

```{r}
raster_stats(mask(chm, chm > 1e6, maskvalues = FALSE), verbose = FALSE)
```

Statistics describe cell values, not area. On a categorical layer
they summarise the class codes, which is rarely meaningful — use
`extract_proportion()` for class composition instead.

## `plot_raster()`

```{r fig.height = 6}
plot_raster(chm)
```

Three defaults are doing work there.

**The colour scale is clamped to central quantiles.** Forty cells
at 250 would otherwise compress everything real into the bottom
few percent of the ramp. Compare the unstretched version, which is
what a naive plot of this layer gives you:

```{r fig.height = 6}
plot_raster(chm, stretch = NULL, main = "No stretch")
```

The pattern is gone. `stretch` takes any two probabilities, and
values beyond them are drawn in the end colours rather than dropped:

```{r fig.height = 6}
plot_raster(chm, stretch = c(0.25, 0.75), main = "Aggressive stretch")
```

**The Alberta boundary is drawn on top**, which is how you see that
a layer is clipped to the province. Pass any polygon to check a
different footprint, or `NULL` for none:

```{r fig.height = 6}
plot_raster(chm, boundary = NULL, main = "No overlay")
```

**`NA` cells are left unpainted.** Giving them a colour turns the
map into a missingness check, which is what finds the hole:

```{r fig.height = 6}
plot_raster(chm, na_col = "firebrick2", main = "Missing cells in red")
```

The gap is obvious, and everything outside the province is flagged
too — masked cells and missing cells are the same thing to a
raster, which is why `raster_stats()` counts them together.

### Multi-layer rasters

Layers are facetted, sharing one colour scale:

```{r fig.height = 4.5}
stack <- c(chm, chm * 1.4)
names(stack) <- c("chm_2020", "chm_2024")
plot_raster(stack, boundary = NULL)
```

A shared scale is what makes two dates comparable, but it is wrong
for bands in different units — an elevation layer beside a slope
layer would flatten one of them. Plot those one at a time
(`x[[1]]`), each getting its own stretch.

### It is a ggplot

Nothing is drawn until the object is printed, so the map composes
and saves like any other plot:

```{r fig.height = 6, message = FALSE, warning = FALSE}
library(ggplot2)

plot_raster(chm, legend_title = "Metres") +
  labs(
    title    = "Canopy height",
    subtitle = "1 km, NAD83 / Alberta 10-TM",
    caption  = "Synthetic layer built in this vignette"
  )
```

```{r eval = FALSE}
p <- plot_raster(chm)
ggsave("2_pipeline/canopy_height.png", p, width = 9, height = 6,
       dpi = 200)
```

Adding a `scale_fill_*()` or a theme replaces the defaults in the
usual way.

## `plot_hist()`

The distribution behind the map, filled along the same ramp so the
two read together:

```{r}
plot_hist(chm)
```

The x axis runs to 250 because the glitch cells are still in there
— the histogram shows the tail the map's stretch hides, which is
the division of labour between them. Clamp them off and the shape
appears:

```{r}
q98 <- raster_stats(chm, quantiles = 0.98, verbose = FALSE)$q98
plot_hist(
  clamp(chm, upper = q98, values = FALSE),
  bins = 80,
  main = "Below the 98th percentile"
)
```

Multi-layer rasters facet here too, but with free scales, since
bands rarely share a value or a count range:

```{r fig.height = 4}
plot_hist(stack)
```

Each panel ramps through the whole palette. The fill is the bin's
position within *its own layer's* range, not an absolute value —
with a shared scale, a layer spanning 0–20 next to one spanning
0–28 would come out two flat blocks of colour. The legend is off by
default for the same reason: it repeats the x axis.

## Theme and palette

Both plots use `theme_science_map()` and the "Hiroshige" palette
from [MetBrewer](https://github.com/BlakeRMills/MetBrewer),
reversed so low values are dark blue and high values red. The ten
hex codes are reproduced inside the package, so MetBrewer is not a
dependency. The theme is exported for other figures in a project:

```{r fig.height = 4}
ggplot(stats, aes(x = layer, y = mean)) +
  geom_col(fill = "#376795", width = 0.4) +
  labs(title = "Any figure, same look", x = NULL, y = "Mean") +
  theme_science_map()
```

Override either by passing `col` or by adding your own scale or
theme to the returned plot.

## A QA pass

The three together, as they would appear after an export finishes:

```{r fig.height = 4.5}
# 1. Does the geometry match what everything else is on?
check_alignment(chm)

# 2. What is in it, and how much of it is missing?
qa <- raster_stats(chm, quantiles = c(0.02, 0.98), verbose = FALSE)
qa[, c("layer", "n_valid", "pct_na", "min", "q98", "max")]

# 3. Is the extreme value a real tail or a handful of cells?
plot_hist(chm, bins = 80)
```

```{r fig.height = 6}
# 4. Does the pattern look like the thing it claims to measure,
#    and is it clipped to Alberta?
plot_raster(chm, na_col = "firebrick2")
```

Which here says: geometry fine, a fifth of the rectangle outside
the province as expected, a gap that needs explaining, and forty
cells that need removing before this becomes a covariate.

## Known gaps

Both plot functions downsample to `maxcell` cells before building
the data frame they draw from, so a province-wide map is a preview
rather than a rendering of every cell, and `plot_hist()` is a sample
of a large layer rather than a census of it. The counts in
`raster_stats()` are exact regardless; only its `quantiles` columns
sample.

Facetted maps share one colour scale, which is wrong for bands in
different units. There is no per-layer stretch — plot single layers
for that.

Neither plot function accepts a directory the way `raster_stats()`
does; they take one raster at a time.
