---
title: "Inspecting layers before you use them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inspecting layers before you use them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Every chunk here is evaluated. The reference layers ship with the
# package and the demonstration layer is built in the vignette, so
# nothing reaches the network or the ABMI share.
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  dpi        = 96
)
```

```{r libraries}
library(sciSpatialR)
library(terra)
```

## Three questions about a layer

A raster that has been pre-processed and aligned with the reference
grid still has to be looked at before it is trusted. Three questions
cover most of what goes wrong, and one function answers each:

1. **What is actually in it?** `raster_stats()`: cell counts,
   missing values, and value summaries as a table.
2. **How are the values distributed?** `plot_hist()`: where they
   pile up, and how long the tails are.
3. **Does the map look right?** `plot_raster()`: the spatial
   pattern, and whether the layer is clipped where it should be.

All three accept a `SpatRaster` or a path, so they work on a layer
in memory and on a freshly exported GeoTIFF.

## A layer to inspect

Everything below runs on a synthetic surface on the reference grid,
built so this vignette needs no external data. It is deliberately
imperfect: a handful of absurd outliers, and a rectangular hole
where an upstream process dropped cells.

```{r build-layer}
set.seed(42)

ref   <- ab_grid()
east  <- init(ref, "x")
north <- init(ref, "y")

# A smooth surface with some structure to look at
chm <- 14 +
  6 * sin(east / 70000) * cos(north / 110000) -
  4 * (north - 5.4e6) / 1.2e6
chm <- chm + init(ref, function(n) rnorm(n, 0, 0.6))

# Sensor glitches: a few cells with impossible values
glitch <- sample(which(!is.na(values(ref, mat = FALSE))), 40)
chm[glitch] <- 250

# A gap left by an upstream process
chm[300:360, 180:240] <- NA

chm <- mask(chm, ref)
names(chm) <- "canopy_height"
chm
```

## `raster_stats()`

```{r stats-default}
raster_stats(chm)
```

Every column is defined above the table. That printing is the
`verbose` argument, on by default; turn it off when the table is
going somewhere other than a human:

```{r stats-quiet}
stats <- raster_stats(chm, verbose = FALSE)
stats$pct_na
```

Quantiles are optional:

```{r stats-quantiles}
raster_stats(chm, quantiles = c(0.02, 0.5, 0.98), verbose = FALSE)
```

### A folder of exports

`x` can also be a path, or a directory, which is the point of the
`source` column: a QA pass over everything that came back from a
batch export is one call.

```{r stats-directory}
export_dir <- file.path(tempdir(), "exports")
dir.create(export_dir, showWarnings = FALSE)

writeRaster(
  aggregate(chm, 8, na.rm = TRUE),
  file.path(export_dir, "canopy_height_8km.tif"),
  overwrite = TRUE
)
writeRaster(
  aggregate(chm, 16, fun = "max", na.rm = TRUE),
  file.path(export_dir, "canopy_max_16km.tif"),
  overwrite = TRUE
)

raster_stats(export_dir, verbose = FALSE)
```

Both rows carry the same layer name, because both files inherited
the band name from the raster they were written from; `source` is
what tells them apart. It is `NA` for a raster held in memory. A
band written *without* a name would come back as terra's `lyr.1`.

An empty layer returns `NA` for the value summaries rather than the
`NaN` and `Inf` that `min()` and `mean()` produce on nothing:

```{r stats-empty}
raster_stats(mask(chm, chm > 1e6, maskvalues = FALSE), verbose = FALSE)
```

Statistics describe cell values. On a categorical layer they
summarise the class codes, which is rarely useful.

## `plot_raster()`

```{r map-default, fig.height = 6}
#| fig.alt: >
#|   Map of the synthetic canopy height layer over Alberta under
#|   the default quantile stretch. A smooth wave pattern runs
#|   across the province, broken by a rectangular blank gap, with
#|   the provincial boundary drawn on top.
plot_raster(chm)
```

Three defaults are doing work there.

**The colour scale is clamped to central quantiles.** A bunch of
cells at an extreme would otherwise compress everything real into
the last few percent of the ramp. Compare the unstretched version,
which is what a naive plot of this layer gives you:

```{r map-no-stretch, fig.height = 6}
#| fig.alt: >
#|   The same layer drawn with no stretch. The forty outlier cells
#|   at 250 metres take over the colour ramp, so the province
#|   renders as one flat colour and the pattern of the previous
#|   map is invisible.
plot_raster(chm, stretch = NULL, main = "No stretch")
```

The pattern is gone. `stretch` takes any two probabilities, and
values beyond them are drawn in the end colours rather than dropped:

```{r map-aggressive-stretch, fig.height = 6}
#| fig.alt: >
#|   The same layer with the colour scale clamped to the 25th and
#|   75th percentiles. The middle of the range is exaggerated into
#|   strong bands and everything beyond the quartiles flattens
#|   into the two end colours.
plot_raster(chm, stretch = c(0.25, 0.75), main = "Aggressive stretch")
```

**The Alberta boundary is drawn on top**, which is how you see that
a layer is clipped to the province. Pass any polygon to check a
different footprint, or `NULL` for none:

```{r map-no-boundary, fig.height = 6}
#| fig.alt: >
#|   The canopy height map with the boundary overlay removed, so
#|   the province's outline is described only by where the layer
#|   has data.
plot_raster(chm, boundary = NULL, main = "No overlay")
```

**`NA` cells are left unpainted.** Giving them a colour turns the
map into a missingness check, which is what finds the hole:

```{r map-missing-cells, fig.height = 6}
#| fig.alt: >
#|   The canopy height map with missing cells painted red, which
#|   picks out both the rectangular gap inside the province and
#|   the whole of the surrounding rectangle outside it.
plot_raster(chm, na_col = "firebrick2", main = "Missing cells in red")
```

The gap is obvious, and everything outside the province is flagged
too.

### Multi-layer rasters

Layers are facetted, sharing one colour scale:

```{r map-multilayer, fig.height = 4.5}
#| fig.alt: >
#|   Two maps side by side, chm_2020 and chm_2024, sharing one
#|   colour scale. The 2024 panel reads uniformly higher on the
#|   ramp because its values are 1.4 times the 2020 panel's.
stack <- c(chm, chm * 1.4)
names(stack) <- c("chm_2020", "chm_2024")
plot_raster(stack, boundary = NULL)
```

A shared scale is what makes two dates comparable, but it is wrong
for bands in different units — an elevation layer beside a slope
layer would flatten one of them. Plot those one at a time
(`x[[1]]`), so they each get their own stretch.

### It is a ggplot

Nothing is drawn until the object is printed, so the map composes
and saves like any other plot:

```{r map-ggplot, fig.height = 6, message = FALSE, warning = FALSE}
#| fig.alt: >
#|   The canopy height map with a ggplot2 title, subtitle and
#|   caption added and the legend relabelled "Metres",
#|   demonstrating that the returned object composes like any
#|   other ggplot.
library(ggplot2)

plot_raster(chm, legend_title = "Metres") +
  labs(
    title    = "Canopy height",
    subtitle = "1 km, NAD83 / Alberta 10-TM",
    caption  = "Synthetic layer built in this vignette"
  )
```

```{r map-save, eval = FALSE}
p <- plot_raster(chm)
ggsave("2_pipeline/canopy_height.png", p, width = 9, height = 6,
       dpi = 200)
```

Adding a `scale_fill_*()` or a theme replaces the defaults in the
usual way.

## `plot_hist()`

The distribution behind the map, filled along the same ramp so the
two read together:

```{r hist-default}
#| fig.alt: >
#|   Histogram of the canopy height cell values, bars filled along
#|   the same colour ramp as the maps. Nearly all the mass sits in
#|   one peak at the low end while the x axis runs out to 250,
#|   exposing the tail the glitch cells create.
plot_hist(chm)
```

The x axis runs to 250 because the glitch cells are still in there
— the histogram shows the tail the map's stretch hides, which is
the division of labour between them. Clamp them off and the shape
appears:

```{r hist-clamped}
#| fig.alt: >
#|   Histogram of the same layer clamped below its 98th
#|   percentile. With the outliers dropped the x axis covers only
#|   the plausible range and the distribution's actual shape
#|   appears.
q98 <- raster_stats(chm, quantiles = 0.98, verbose = FALSE)$q98
plot_hist(
  clamp(chm, upper = q98, values = FALSE),
  bins = 80,
  main = "Below the 98th percentile"
)
```

Multi-layer rasters facet here too, but with free scales, since
bands rarely share a value or a count range:

```{r hist-multilayer, fig.height = 4}
#| fig.alt: >
#|   Two histograms side by side, chm_2020 and chm_2024, each on
#|   its own free axis scale, with both panels ramping through the
#|   full palette.
plot_hist(stack)
```

Each panel ramps through the whole palette. The fill is the bin's
position within its own layer's range, not an absolute value.

## Theme and palette

Both plots use the "Hiroshige" palette from
[MetBrewer](https://github.com/BlakeRMills/MetBrewer), reversed so
low values are dark blue and high values red. The ten hex codes are
reproduced inside the package, so MetBrewer is not a dependency. The
theme is exported for other figures in a project
(`theme_science_map()`).

```{r theme-reuse, fig.height = 4}
#| fig.alt: >
#|   A column chart of the layer's mean value drawn as a single
#|   solid blue bar with theme_science_map() applied, showing the
#|   package theme carried onto a figure that is not a map.
ggplot(stats, aes(x = layer, y = mean)) +
  geom_col(fill = "#376795", width = 0.4) +
  labs(title = "Any figure, same look", x = NULL, y = "Mean") +
  theme_science_map()
```

Override either by passing `col` or by adding your own scale or
theme to the returned plot.

## Known gaps

Both plot functions downsample to `maxcell` (5e5) cells before
building the data frame they draw from, so a province-wide map is a
preview rather than a rendering of every cell, and `plot_hist()` is
a sample of a large layer rather than a census of it. The counts in
`raster_stats()` are exact regardless; only its `quantiles` columns
sample.

Facetted maps share one colour scale, which is wrong for bands in
different units. There is no per-layer stretch — plot single layers
for that.

Neither plot function accepts a directory the way `raster_stats()`
does; they take one raster at a time.
